{ Das CrypTool-Skript: \\ Kryptographie, Mathematik und mehr} \\ \vspace{30pt} {\large Hintergrundmaterial und Zusatzinformationen\\%*[-8pt] zum freien E-Learning-Programm CrypTool\\ (9. Auflage -- zu Version 1.4.20)} \vspace{60pt}

### 2009-07-20 script/de/chapters/paper_and_pencil.tex
CT script, paper and pencil: some small changes in de and en; translated chapter 2.5 from en to de; font of cover page adapted

Das folgende Kapitel bietet einen recht vollständigen Überblick über Papier- 
und Bleistiftverfahren\footnote{Jeweils mit Verweisen zu vertiefenden Informationen.}.
Unter diesem Begriff lassen sich alle Verfahren zusammenfassen, die Menschen von
Hand anwenden können, um Nachrichten zu ver- und entschlüsseln.
Besonders populär waren diese Verfahren für Geheimdienste (und 
sind es immer noch), da ein Schreibblock und ein Stift -- im 
Gegensatz zu elektronischen Hilfsmitteln -- vollkommen unverdächtig sind.

@@

\vskip +30 pt
%------------------------------------------------------------------------------
\section{Transpositionsverfahren}

@@

%------------------------------------------------------------------------------
\section{Substitutionsverfahren}
\index{Substitution}

@@

%------------------------------------------------------------------------------
\section{Kombination aus Substitution und Transposition}

@@

% ---------------------------------------------------------------------------
\newpage
\section{Beispiele mit Sage}
\index{Sage}

@@

Eine erste Einführung in das CAS Sage finden Sie im Anhang \ref{s:appendix-using-sage}.

Um die Beispielprogramme leichter lesbar zu machen, hielten wir uns an die in der folgenden
Grafik aufgeführten Bezeichnungen und Abläufe:
\begin{itemize}
  \item Die Verschlüsselung besteht aus den Schritten codieren und verschlüsseln.
  \item Beim Codieren werden in dem gegebenen Klartext P die Zeichen auf Groß- bzw.
        Kleinbuchstaben angepasst (je nach gegebenem Alphabet), und alle
	Nichtalphabetzeichen werden entfernt.
  \item Durch die Verschlüsselung wird das Chiffrat C erzeugt.
  \item Die Entschlüsselung besteht ebenfalls aus den zwei Schritten entschlüsseln und decodieren.
  \item Decodieren ist nur nötig, wenn die benutzten Symbole im Alphabet nicht die ASCII-Zeichen
        sind ????????????????.
\end{itemize}

@@

\newpage
\subsection{Transposition ciphers}

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build the spave of our plaintext
and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by two characters, with wrap
around. That is the encryption process. The decryption process is
shifting each letter of the ciphertext to the left by $14 - 2 = 12$
characters in order to recover the plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # given plaintext
sage: P   = "a b c d e f g h i j k l m n"
sage: # encryption key
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars, convert lower-case to upper-case)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in two steps)
sage: E   = T(key)
sage: C   = E(msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D   = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
sage:
sage: # Representation of key and inverse key as permutations
sage: E
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: D
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key explicitely given)}
\end{sagecode}

\newpage
The next sample is also a simple column transposition. But now the code is a little bit more automated in generating the keys from the shift parameter.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key constructed with "range")}
\end{sagecode}

\newpage
As a more complicated example, we can use a permutation in the
encryption and decryption processes. If the block length of our
transposition cipher is $n$, then the encryption/decryption
key must be a permutation on $n$ symbols. The following
example uses the method \verb!random_key()! of the class
\verb!TranspositionCryptosystem!. So each call to
\verb!random_key()! produces a different key for encryption, which
affects the ciphertext. Note that your results may be different from
the following example.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # Remark: Enciphering here requires, that the length of msg is a multiple of keylen
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,3,13,6,5,4,12,7)(11,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage: C   = T.enciphering(key, msg); C
BCMLDEAHIJNGFKPQAZRSOVWXBUTY
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
ssage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Just another way of decryption: Using "enciphering" with the inverse key
sage: keyInv = T.inverse_key(key); keyInv
(1,7,12,4,5,6,13,3,2)(11,14)
sage: DC     = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Test correctness of decryption
sage: msg == DC
True
\end{Verbatim}
\caption{Simple Column Transposition with randomly generated (permutation) key}
\end{sagecode}

\newpage
The following example shows the key space of a transpositon cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage: T.key_space()
Symmetric group of order 14! as a permutation group
sage: # Remark: The key space is not quite correct as also permutations shorter than keylen are counted.
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # enciphering in one and in two steps
sage: C   = T.enciphering(key, msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: enc = T(key); enc.key()
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: C = enc(msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: # deciphering
sage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
\end{Verbatim}
\caption{Simple Column Transposition (showing the key\_space)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitution ciphers}

Substitution cryptosystems are implemented in Sage in the class

\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent The following code sample uses Sage to construct a substitution
cipher with a random key. A random key can be generated using the
method \verb!random_key()! of the class
\texttt{SubstitutionCrypto\-system}. Different keys determine different
substitution ciphers. So each call to \verb!random_key()! returns
different results. First, we perform the encryption process with a
random key.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #.# We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Caesar cipher}

The following example uses Sage to construct a Caesar
cipher. First, we demonstrate encryption using this cipher.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitely given, step-by-step approach)}
\end{sagecode}

\newpage
The next sample does the same, but the code is more sophisticated/automated/variable.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; keys generated)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitution with symbols}

Here is a Sage example where the symbols are from the binary number
system. A monoalphabetic substitution cipher with a binary alphabet has very little
security: Because the plaintext/ciphertext alphabet has only the two elements
0 and 1, there are only two keys possible: (0 1) and (1 0).

>>>>

Das folgende Kapitel bietet einen recht vollständigen Überblick über Papier- 
und Bleistiftverfahren\footnote{%
In den Fußnoten zu diesem Kapitel wird jeweils
beschrieben, wie man diese Verfahren mit CrypTool 1 ausführen kann.
Außerdem enthält das letzte Unterkapitel (\ref{PaP_Sage_samples})
zu einigen Verfahren Beispielcode in dem Computer-Algebra-System Sage\index{Sage}.
                               }.
Unter diesem Begriff lassen sich alle Verfahren zusammenfassen, die Menschen von
Hand anwenden können, um Nachrichten zu ver- und entschlüsseln.
Besonders populär waren diese Verfahren für Geheimdienste (und 
sind es immer noch), da ein Schreibblock und ein Stift -- im 
Gegensatz zu elektronischen Hilfsmitteln -- vollkommen unverdächtig sind.

@@

\newpage
%------------------------------------------------------------------------------
\section{Transpositionsverfahren}
\label{PaP_transposition_ciphers}
\index{Transposition}

@@

%------------------------------------------------------------------------------
\section{Substitutionsverfahren}
\label{PaP_substitution_ciphers}
\index{Substitution}

@@

%------------------------------------------------------------------------------
\section{Kombination aus Substitution und Transposition}
\label{PaP_combined_ciphers}
\index{Kaskaden}

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\section{Programmbeispiele mit Sage}
\label{PaP_Sage_samples}
\index{Sage!Code examples}
\index{Sage}

@@

\noindent Eine erste Einführung in das CAS Sage finden Sie im Anhang \ref{s:appendix-using-sage}.

\noindent Um die Beispielprogramme\footnote{%
     Weitere Programmbeispiele in Sage finden sich z.B. unter:\\
     \href{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}{\tt
           http://www.sagemath.org/doc/constructions/linear\_codes.html\#classical-ciphers}
                                          }
leichter lesbar zu machen, hielten wir uns an die in der folgenden
Grafik aufgeführten Bezeichnungen und Abläufe:
\begin{itemize}
  \item Die Verschlüsselung besteht aus den Schritten codieren und verschlüsseln.
  \item Beim Codieren werden in dem gegebenen Klartext P die Zeichen auf Groß- bzw.
        Kleinbuchstaben angepasst (je nach gegebenem Alphabet), und alle
	Nichtalphabetzeichen werden entfernt.
  \item Durch die Verschlüsselung wird das Chiffrat C erzeugt.
  \item Die Entschlüsselung besteht ebenfalls aus den zwei Schritten entschlüsseln und decodieren.
  \item Decodieren ist nur nötig, wenn die benutzten Symbole im Alphabet nicht die ASCII-Zeichen
        sind.
\end{itemize}

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Transpositions-Chiffren}

Tranpositions-Chiffren sind implementiert in der Sage-Klasse
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
Bevor man mit einer Sage Transpositions-Chiffre arbeiten kann, muss man sie
konstruieren: Dazu legt man das Alphabet fest, das sie Symbole (Zeichen) enthält,
aus denen Klartext und Geheimtext bestehen können. 
Typischerweise besteht das Alphabet aus den normalen lateinischen Großbuchstaben.
Dieses Alphabet legt man mit der folgenden Funktion fest
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
Danach muss man die Blocklänge der Permutation festlegen, d.h. die
Länge des Zeilenvektors der einfachen Spaltentransposition.
Dieser Zeilenvektor ist der Schlüssel, mit dem die Buchstaben des
Klartextes permutiert werden.

Im folgenden ersten Beispiel der Transpositions-Chiffren ist die 
Blocklänge 14, und der Schlüssel ist so gebaut, dass jeder Buchstabe
im Klartext um zwei Zeichen nach rechts geshiftet wird (und wrap-around
am Ende des Blocks). Das ist der Verschlüsselungsvorgang.
Der Entschlüsselungsvorgang besteht im Shiften jedes Zeichens des
Geheimtextes um $14 - 2 = 12$ Zeichen nach links.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # given plaintext
sage: P   = "a b c d e f g h i j k l m n"
sage: # encryption key
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars, convert lower-case to upper-case)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in two steps)
sage: E   = T(key)
sage: C   = E(msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D   = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
sage:
sage: # Representation of key and inverse key as permutations
sage: E
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: D
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Einfache Transposition durch Shiften (die Schlüssel sind explizit gegeben)}
\end{sagecode}

\newpage
Das zweite Beispiel der Transpositions-Chiffren ist ebenfalls eine einfache
shiftende Spaltentransposition. Aber hier ist der Programmcode ein wenig mehr
automatisiert: Die Schlüssel werden anhand des Shift-Wertes generiert.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Einfache Transposition durch Shiften (die Schlüssel werden mit "range" konstruiert)}
\end{sagecode}

\newpage
Im dritten Beispiel der Transpositions-Chiffren wird eine beliebige Permutation
als Schlüssel für die Ver- und Entschlüsselun gewählt, um die Zeichen in jedem Block
zu verwürfeln (Blocklänge = Anzahl Spalten in der einfachen Spaltentransposition).
Ist die Blocklänge $n$, dann ist der Schlüssel eine Permutation über $n$ Zeichen.
Das folgende Beispiel nutzt die Methode \verb!random_key()! der Klasse
\verb!TranspositionCryptosystem!.  Jeder Aufruf von \verb!random_key()! erzeugt
einen anderen Schlüssel. Deshalb werden bei Ihrem Aufruf voraussichtlich andere
Ergebnisse (Schlüssel und Geheimtext) auftreten.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # Remark: Enciphering here requires, that the length of msg is a multiple of keylen
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,3,13,6,5,4,12,7)(11,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage: C   = T.enciphering(key, msg); C
BCMLDEAHIJNGFKPQAZRSOVWXBUTY
sage: # decryption using the "deciphering method with key" instead of "enciphering with keyInv" 
ssage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Just another way of decryption: Using "enciphering" with the inverse key
sage: keyInv = T.inverse_key(key); keyInv
(1,7,12,4,5,6,13,3,2)(11,14)
sage: DC     = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Test correctness of decryption
sage: msg == DC
True
\end{Verbatim}
\caption{Einfache Spalten-Transposition mit zufällig erzeugtem (Permutations-)Schlüssel}
\end{sagecode}

\newpage
Das vierte Beispiel der Transpositions-Chiffren gibt zusätzlich die Größe des Schlüsselraums
einer einfachen Spaltentransposition aus.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage: T.key_space()
Symmetric group of order 14! as a permutation group
sage: # Remark: The key space is not quite correct as also permutations shorter than keylen are counted.
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # enciphering in one and in two steps
sage: C   = T.enciphering(key, msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: enc = T(key); enc.key()
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: C = enc(msg); C
BGIEJNAMCLDHKFPUWSXBOAQZRVYT
sage:
sage: # deciphering
sage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
\end{Verbatim}
\caption{Einfache Spalten-Transposition (mit Ausgabe der Größe des Schlüsselraumes)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitutions-Chiffren}

Substitutions-Verschlüsselungen sind in Sage in der folgenden Klasse implementiert
\begin{center}
\verb!sage.crypto.classical.SubstitutionCryptosystem!
\end{center}

\noindent Das folgende Programmbeispiel lässt Sage eine Substitutions-Chiffre mit
zufälligem Schlüssel erstellen. Den zufälligen Schlüssel kann man mit der Methode
\verb!random_key()! der Klasse \texttt{SubstitutionCrypto\-system} erzeugen.
Unterschiedliche Schlüssel erzeugen eine unterschiedliche Substitutions-Chiffre.
Deshalb erhält man mit jedem Aufruf von \verb!random_key()! voraussichtlich
ein anderes Ergebnis.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #### We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetische Substitution mit zufällig erzeugtem Schlüssel}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Caesar-Chiffre}

Das folgende Programmbeispiel erzeugt eine Caesar-Chiffre.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #### Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (Substitution durch Shiften des Alphabets; Schlüssel explizit gegeben; Schritt-für-Schritt-Ansatz)}
\end{sagecode}

\newpage
Das zweite Caesar-Beispiel macht dasselbe, aber der Programmcode ist variabler.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: S
Substitution cryptosystem on Free alphabetic string monoid on A-Z
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: len(key)
26
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #### Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (Substitution durch Shiften des Alphabets; Substitutions-Schlüssel werden berechnet)}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitutions-Chiffre mit Symbolen}

Im folgenden Sage-Programmbeispiel sind die Alphabetzeichen aus dem Binärsystem.
Eine monoalphabetische Substitution über einem Binäralphabet bietet nur wenig
Sicherheit: Weil das Klartext-/Geheintext-Alphabet nur aus den 2 Elementen 
0 and 1 besteht, gibt es nur zwei mögliche Schlüssel: (0 1) and (1 0).
Anmerkung: In jedem Schlüssel einer Substitutions-Chiffre müssen alle
Alphabetzeihen genau einmal auftreten.
